Well-quasi-orderings on word languages

### 2024-08-06 src/prefixes.tex
adding some proofs

\begin{example}
    The set $L = \setof{a^nb}{n \in \Nat}$ is an infinite antichain for the
    prefix relation.
\end{example}

@@

@@

\begin{definition}
    An \intro{antichain branch} for a language $L$ is an infinite 
    branch $B$ of the tree $T$ such that from every point of the branch, 
    one can reach a word of $A$ that is not in the branch.
\end{definition}

@@

\begin{lemma}
    Let $L \subseteq \Sigma^*$ be a language. Then, $L$ contains an infinite
    \kl{antichain} if and only if there exists an \kl{antichain branch} for $L$.
\end{lemma}

\begin{corollary}
    If $L$ is regular, then the existence of an infinite antichain is decidable.
\end{corollary}

@@

\begin{theorem}
    A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the
    \kl{prefix relation} if and only if $L$ is a union of \kl{chains}.
\end{theorem}

>>>>

\begin{example}
    The set $L = \setof{a^nb}{n \in \Nat} \subseteq \set{a,b}^*$ is an infinite \kl{antichain} for the
    \kl{prefix relation}.
\end{example}

@@

\begin{figure}
    \centering
    \caption{The tree of prefixes over the alphabet $\set{a,b}$,
        with an \kl{antichain} in red, and
        an \kl{antichain branch} in blue.}
\end{figure}

@@

\begin{definition}
    An \intro{antichain branch} for a language $L$ is an infinite 
    branch $B$ of the \kl{tree of prefixes} such that from every point of the branch, 
    one can reach a word of $L$ that is not in the branch.
\end{definition}

@@

\begin{lemma}
    Let $L \subseteq \Sigma^*$ be a language. Then, $L$ contains an infinite
    \kl{antichain} if and only if there exists an \kl{antichain branch} for $L$.
\end{lemma}
\begin{proof}
    Assume that $L$ contains an \kl{antichain branch}. Let us construct an
    infinite \kl{antichain} as follows. We start with a set $A_0 \defined
    \emptyset$ and a node $v_0$ at the root of the tree. At step $i$, we
    consider a word $w_i$ such that $v_i$ is a \kl{prefix} of $w_i$, and $w_i
    \in L \setminus B$, which exists by definition of \kl{antichain branches}.
    We then set $A_{i+1} \defined A_i \cup \set{w_i}$. To compute $v_{i+1}$, we
    consider the largest prefix of $w_i$ that belongs to $B$, and set $v_{i+1}$
    to be the successor of this prefix in $B$. By an immediate induction, we
    conclude that for all $i \in \Nat$, $A_i$ is an \kl{antichain}, and that
    $v_i$ is a node in the \kl{antichain branch} $B$ such that $v_i$ is not a
    prefix of any word in $A_i$.

    Conversely, assume that $L$ contains an infinite \kl{antichain} $A$. Let us
    construct an \kl{antichain branch}. Let us consider the subtree of the
    \kl{tree of prefixes} that consists in words that are \kl{prefixes} of
    words in $A$. This subtree is infinite, and by König's lemma, it contains
    an infinite branch. By definition this is an \kl{antichain branch}.
\end{proof}

\begin{corollary}
    If $L$ is regular, then the existence of an infinite antichain is decidable.
\end{corollary}
\begin{proof}
    If $L$ is regular, then it is \kl{$\MSO$-definable}, and there 
    exists a formula $\varphi(x)$ in \kl{$\MSO$} that selects nodes 
    of the \kl{tree of prefixes} that belong to $L$. Now, to decide whether there
    exists an \kl{antichain branch} for $L$, we can simply check whether
    the following formula is satisfied:
    \begin{equation*}
        \exists B. 
        B \text{ is a branch } \land
        \forall x \in B, \exists y. y \text{ is a child of } x \land \varphi(y) \land y \not\in B
        \quad .
    \end{equation*}
    Because the above formula is an $\MSO$-formula over the infinite
    binary tree, whether it is satisfied is decidable.
    \textbf{TODO: cite}.
\end{proof}

@@

\begin{theorem}
    A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the
    \kl{prefix relation} if and only if $L$ is a union of \kl{chains}.
\end{theorem}
\begin{proof}
    Assume that $L$ is a finite union of \kl{chains}.
    Because the \kl{prefix relation} is \kl{well-founded},
    and that finite unions of \kl{chains} have finite \kl{antichains},
    we conclude that $L$ is \kl{well-quasi-ordered}.

    Conversely, assume that $L$ is not a finite union of \kl{chains}. Then
    there exists an infinite \kl{antichain branch} $B$ for $L$. Indeed
    \textbf{todo}. In particular, $L$ is not \kl{well-quasi-ordered}.
\end{proof}
